\documentclass[12pt,a4paper]{report} % or your university's class

% ---------- Encoding & language ----------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}

% ---------- Typography & layout ----------
\usepackage{lmodern}
\usepackage{setspace}
\onehalfspacing
\usepackage{geometry}
\geometry{margin=2.5cm}

% ---------- Math / equations ----------
\usepackage{amsmath,amssymb,amsthm,mathtools}

% ---------- Tables ----------
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{siunitx}

% ---------- Figures ----------
\usepackage{graphicx}
\usepackage{float}
\usepackage{subcaption}
\usepackage{caption}

% ---------- References / hyperlinks ----------
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}
\usepackage{csquotes}

% ---------- Bibliography (biblatex + biber) ----------
\usepackage[
    backend=biber,
    style=ieee,      % Uses numeric citations [1] and IEEE bibliography format
    sorting=none     % CRITICAL: Sorts by order of appearance (Standard IEEE)
]{biblatex}
\addbibresource{reference.bib}

% ---------- Other useful packages ----------
\usepackage{enumitem}
\usepackage{xcolor}

\begin{document}

% ---------- Front matter ----------
\title{Trustworthy RAG: An Evaluation Agent for Detecting Misinformation and Knowledge Poisoning in Generative AI Systems}
\author{Balkrishna Giri}
\date{\today} % or fixed date

\maketitle

\begin{abstract}
Retrieval-Augmented Generation (RAG) systems improve large language model reliability by grounding responses in external knowledge. However, RAG architectures inherently trust their retrieval output, creating a critical Security-Reliability Gap: high semantic relevance does not guarantee factual truth. Adversaries can exploit this by injecting malicious documents into the knowledge corpus --- a threat known as Knowledge Poisoning --- causing the system to generate targeted misinformation.

This thesis proposes an autonomous Evaluation Agent as a defensive middleware within the RAG pipeline. The agent combines Natural Language Inference (NLI) for factual verification, a multi-signal Poison Detector (linguistic, structural, intra-document, cross-document, and semantic-outlier signals), and a composite Trust Index ($T = 0.4 \cdot F + 0.35 \cdot C + 0.25 \cdot (1-P)$) to produce an interpretable trustworthiness score for each generated response.

Experiments on TruthfulQA and FEVER benchmarks with four adversarial poisoning strategies (contradiction, instruction injection, entity swap, and subtle manipulation) demonstrate that the system achieves 91\% accuracy and 57.1\% F1 score with 100\% precision (zero false positives) using Llama~3.3~70B, a +7\% improvement over the naive always-trust baseline. Instruction injection is detected with 100\% recall, while entity-swap and subtle manipulation remain near-undetectable at the textual level, representing an architectural limit of surface-signal-based approaches.

A 2×2 factorial experiment (\{Llama~3.3~70B, Qwen~3.5~35B\} × \{all-MiniLM-L6-v2, Snowflake Arctic Embed2\}) reveals that the Trust Index is LLM-dependent: Qwen~3.5~35B's verbose generation style causes systematic false positives, underperforming the naive baseline by 16\%, while LLM selection dominates embedding model choice as the key performance factor. A retrieval depth sensitivity analysis ($K=3$ vs.\ $K=5$) confirms that detection performance is retrieval-depth-invariant: the 40\% recall ceiling is set by attack strategy difficulty, not by the number of documents retrieved.

The work addresses a gap left by existing frameworks such as RAGAS and ARES, which measure faithfulness to context but not the integrity of the context itself.
\end{abstract}

\tableofcontents
\listoffigures      % optional
\listoftables       % optional

% ---------- Main chapters ----------
\include{chapters/ch1_introduction}
\include{chapters/ch2_theory_background}
\include{chapters/ch3_literature_review}
\include{chapters/ch4_methodology}
\include{chapters/ch5_eval_agent}
\include{chapters/ch6_experiments}
\include{chapters/ch7_discussion}
\include{chapters/ch8_conclusion}

% ---------- References ----------
\printbibliography

% ---------- Appendices ----------
\appendix
\include{chapters/appA_parameters}
\include{chapters/appB_results}
\include{chapters/appC_prompts}
\include{chapters/appD_ethics}
\include{chapters/appE_implementation}

\end{document}
